A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-15 tidyindex.qmd
more small tweaks

-   $x^i(s, t)$ represents a variable of interest for example precipitation,$j = 1, \cdots, p$ and

@@

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|--------------|----------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x(s;t)]$                                                                       | $Tx^i_t(s)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x(s; t)]$                                                              | $z(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x(s;t)]$                                                                       | $b(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | ccccc                                                                             | $A(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$ , as the input. It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI).

@@

```{=tex}
\begin{equation}
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}
```
where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category. In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$. In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.

>>>>

-   $x^i(s, t)$ represents a variable of interest for example precipitation, $i = 1, \cdots, p$ and

@@

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|--------------|----------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F_{\mathbf{\theta}}(.)$ is the CDF function | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$, as the input. It summarises the high-dimensional information into a lower dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < q$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided in statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI).

@@

```{=tex}
\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}
```
where $C_0, C_1,\cdots ,C_z$ are the categories and $c_0, c_1, \cdots, c_z$ are the thresholds for each category. In SPI, droughts are sorted into four categories: mild drought: $[-0.99, 0]$; moderate drought: $[-1.49, -1]$; severe drought: $[-1.99, -1.5]$, and extreme drought: $[-\infty, -2]$. In this case, $C_0, C_1, C_2, C_3$ are the drought categories: mild, moderate, severe, and extreme drought ($z = 3$) and $c_0 =0, c_1 = -1, c_2 = -1.5, c_3 = -2$ are the cutoff value for each class.

### 2023-08-15 tidyindex.tex
more small tweaks

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  \(x^i(s, t)\) represents a variable of interest for example
  precipitation,\(j = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season.
\end{itemize}

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x(s;t)]\) &
\(Tx^i_t(s)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x(s;t)]\), where
\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p(s;t)\), where
\(p(.) \in [0, 1]\) is a probability value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x(s; t)]\) & \(z(s; t)\),
where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x(s;t)]\) & \(b(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & ccccc &
\(A(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\) , as
the input. It summarises the high-dimensional information into a lower
dimension representation \(\mathbf{y}(s;t)\), where
\(y \subseteq \mathbb{R}^d\) and \(d < p\), as the output. The
transformation can be based on domain-specific knowledge, originating
from theories describing the underlying physical processes, or guided in
statistical methods. For example, the Standardized
Precipitation-Evapotranspiration Index (SPEI) calculates the difference
\(D\) between precipitation (\(P\)) and potential evapotranspiration
(\(\text{PET}\)), using a water balance model (\(D = P - \text{PET}\)).
This is the only step that differs from the Standardized Precipitation
Index (SPI).

@@

\begin{equation}
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, boxrule=0pt, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden]}{\end{tcolorbox}}\fi

@@

\begin{itemize}
\item
  \(x^i(s, t)\) represents a variable of interest for example
  precipitation, \(i = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season.
\end{itemize}

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1795}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
\(Tx^i(s;t)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where
\(F_{\mathbf{\theta}}(.)\) is the CDF function & \(p^i(s;t)\), where
\(p(.) \in [0, 1]\) is a probability value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
\(z^i(s; t)\), where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\), as the input. It
summarises the high-dimensional information into a lower dimension
representation \(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\)
and \(d < q\), as the output. The transformation can be based on
domain-specific knowledge, originating from theories describing the
underlying physical processes, or guided in statistical methods. For
example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
calculates the difference \(D\) between precipitation (\(P\)) and
potential evapotranspiration (\(\text{PET}\)), using a water balance
model (\(D = P - \text{PET}\)). This is the only step that differs from
the Standardized Precipitation Index (SPI).

@@

\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x(s; t) < C_0 \\
a_1 & C_2 \leq x(s; t) < C_1 \\
a_2 & C_3 \leq x(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x(s; t)
\end{cases}
\end{equation}
